Abstraction and Invariance for Algebraically Indexed Types

### 2012-07-09 paper/section-monoid.tex
Minor spacing changes to the monoid section

\begin{example}
  By selecting a suitable monoid $M$, we can simulate a programming
  language that annotates strings with taint markers. Let $\Sigma$ be
  a set of characters from some alphabet, and let $T$ be a set of
  taint markers. Set $M = \Sigma^* \times
  \mathcal{MS}_{\mathit{fin}}(T)$, where $\mathcal{MS}_{\mathit{fin}}$
  denotes the finite power set operator. Since both $\Sigma^*$ and
  $\mathcal{MS}_{\mathit{fin}}(T)$ are monoids (the latter with
  multiset union as the monoid multiplication, which we write as
  $\uplus$), $M$ is itself a monoid. The centre $Z(M)$ of $M$, 
  consists of all elements of the form $([], t)$.

  As an example of a free theorem for this instantiation of our
  general framework, consider programs of the following type:
  \begin{displaymath}
    ; \Gamma_\Mon \vdash N : \forall a,b\mathord:\mathsf{M}.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{aa}
  \end{displaymath}
  From this type we can derive the following free theorem about
  $N$. For all $a, b \in \mathcal{MS}_{\mathit{fin}}(T)$, $(s_1,t_1),
  (s_2,t_2) \in \Sigma^*$,
  \begin{displaymath}
    \begin{array}{l}
      \pi_2(\tmSem{N}\ (s_1,t_1 \uplus a)\ (s_2,t_2 \uplus b)) = \\
      \hspace{2em}\pi_2(\tmSem{N}\ (s_1,t_1)\ (s_2,t_2)) \uplus a \uplus a
    \end{array}
  \end{displaymath}
  If we taint the first input by $a$ and the second by $b$, then two
  copies of the taint $a$ will flow to the output, and none of taint
  $b$.
\end{example}

@@

The proof of this theorem uses the non-compositional relational
environments from \autoref{sec:constr-rel-env}, with
$R^\bullet_{\tyPrimNm{M}} = \emptyset$.

\begin{example}
  By \thmref{thm:monoid-indefinability}, it is impossible to write a
  program with the type $\tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_1}$, and is thus % shows that it is
  impossible to convert a doubly tainted value to a singly tainted
  one.
%
  By contrast with the situation for abelian group indexed types, the
  lack of inverses in the monoid indexed setting means that the type
  $\tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_2} \to \tyPrim{M}{m_1}$ is also
  uninhabited since there are no naturals $a_1,a_2$ such that
  $(m_1m_2)^{a_1}m_2^{a_2} \equiv m_1$.
\end{example}

>>>>

By selecting a suitable monoid $M$, we can simulate a programming
language that annotates strings with taint markers. Let $\Sigma$ be a
set of characters from some alphabet, and let $T$ be a set of taint
markers. Set $M = \Sigma^* \times \mathcal{MS}_{\mathit{fin}}(T)$,
where $\mathcal{MS}_{\mathit{fin}}$ denotes the finite power set
operator. Since both $\Sigma^*$ and $\mathcal{MS}_{\mathit{fin}}(T)$
are monoids (the latter with multiset union as the monoid
multiplication, which we write as $\uplus$), $M$ is itself a
monoid. The centre $Z(M)$ of $M$, consists of all elements of the form
$([], t)$.

As an example of a free theorem for this instantiation of our general
framework, consider programs of the following type:
\begin{displaymath}
  ; \Gamma_\Mon \vdash N : \forall a,b\mathord:\mathsf{M}.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{aa}
\end{displaymath}
From this type we can derive the following free theorem about $N$. For
all $a, b \in \mathcal{MS}_{\mathit{fin}}(T)$, $(s_1,t_1), (s_2,t_2)
\in \Sigma^*$,
\begin{displaymath}
  \begin{array}{l}
    \pi_2(\tmSem{N}\ (s_1,t_1 \uplus a)\ (s_2,t_2 \uplus b)) = \\
    \hspace{2em}\pi_2(\tmSem{N}\ (s_1,t_1)\ (s_2,t_2)) \uplus a \uplus a
  \end{array}
\end{displaymath}
If we taint the first input by $a$ and the second by $b$, then two
copies of the taint $a$ will flow to the output, and none of taint
$b$.

@@

The proof of this theorem uses the non-compositional relational
environments from \autoref{sec:constr-rel-env}, with
$R^\bullet_{\tyPrimNm{M}} = \emptyset$. From the theorem we can see
that it is impossible to write a program with the type
$\tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_1}$, and is thus impossible to
convert a doubly tainted value to a singly tainted one.
%
By contrast with the situation for abelian group indexed types, the
lack of inverses in the monoid indexed setting means that the type
$\tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_2} \to \tyPrim{M}{m_1}$ is also
uninhabited since there are no naturals $a_1,a_2$ such that
$(m_1m_2)^{a_1}m_2^{a_2} \equiv m_1$.
